\propertitle

### 2019-11-21 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-21T11:47:50+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

This sampling relation is even more straightforward if instead of power
moments we use \emph{normalized factorial moments} \citep{potts1953}. The
$m$th normalized factorial moment of a distribution $\pxx(\ya)$ is
\begin{equation}
  \label{eq:def_factorial_moment_a}
  \E\Biggl[\binom{\ya}{m} \Biggr]\bigg/\binom{n}{m}
  \equiv
  \sum_{\ya=0}^{n} %m!\,
  \displaystyle\binom{\ya}{m} \bigg/ \binom{n}{m}\; \pxx(\ya),
  \qquad 1\le m \le n,
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
that is, the expected number of distinct $m$-tuples of simultaneously
spiking neurons (within a time-bin's width), normalized by the maximum
possible number of distinct $m$-tuples. % For
% example, with $m=2$, if $a=4$ neurons spike in a network of $n=5$, we
% have $\tbinom{4}{2}=6$ distinct pairs of simultaneously spiking neurons,
% and the total number of distinct pairs is $\tbinom{5}{2}=10$. The
% normalized number of spiking pairs is therefore $6/10$.
Note that the first $m$ factorial moments together provide the same
information as the first $m$ power moments together, and vice versa: they
are linearly related because $\tbinom{\ya}{m}$ is a polynomial in $\ya$ of
degree $m$. % Specifying one set is therefore
% equivalent to specifying the other set.
But the normalized factorial moments have a convenient property for our
analysis: \emph{the first $n$ normalized factorial moments for the sample
  and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \begin{split}
  \E\Biggl[\binom{\ya}{m} \Biggr]\bigg/\binom{n}{m}
  &=
  \E\Biggl[\binom{\yA}{m} \Biggr]\bigg/\binom{N}{m}
  \qquad\text{or}\\
    \sum_{\ya=0}^{n} %m!\,
  \binom{\ya}{m}\bigg/  \binom{n}{m}\; \pxx(\ya)
&=    \sum_{\yA=0}^{N} %m!\,
  \binom{\yA}{m}\bigg/  \binom{N}{m}\; \pxxx(\yA),\qquad
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  1\le m \le n.
\end{split}
\end{equation}
% where $\pxxx(\yA)$ is our distribution of belief about the full-network
% activity, and $\pxx$

@@

Before applying the formula above to two concrete data sets we want to add
two remarks on the maximum-entropy method that are seldom made in the
neuroscientific literature. They are important for the interpretation of
the results and are further discussed in \sect~\ref{sec:assumptions}.
First, a maximum-entropy distribution, like $\px(\yA \| M, N)$, is the
zeroth-order approximation (in the sense of Laplace's
approximation)\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000}
of four different distributions for the full network, which differ
numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
The validity of the maximum-entropy approximation decreases as the ratio
$N/T$ increases. Second, the maximum-entropy
method % (based on the \emph{Shannon} relative
% entropy, as opposed to other entropy functions such as Burg's
% \citeyear{burg1975})
rests on some implicit assumptions about the probabilities for the long-run
frequency distribution of activities, besides the assumptions mentioned at
the end of the previous paragraph. So the often heard statement that it
gives \enquote{the maximally unbiased (or non-committal) distribution} must
be taken with a grain of salt. \citep[\mynote{move to
  sect}]{jaynes1986d_r1996,portamana2009,portamana2017}.

@@

In the next section we apply the method just described to the data sets
from two actual recordings, using six moments, and discuss the properties
of the resulting distributions.

@@

We apply the approach just described to two data sets publicly available in
the literature:
\begin{itemize}[wide]
\item The \textbf{first data set}, from \textcite[rat
  14147]{stensolaetal2012}, consists of $n=65$ neurons (27 of which
  classified as grid cells) from rat Medial Entorhinal Cortex, recorded for
  about 20 minutes. Their spikes are binned into $T=417\,641$ bins of
  $3\;\textrm{ms}$ width.
\item The \textbf{second data set}, from
  \textcite{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
  consists of $n=159$ neurons from macaque Motor Cortex, recorded for about
  15 minutes. Their spikes are binned into $T=300\,394$ bins of
  $3\;\textrm{ms}$ width.
\end{itemize}

@@

We first calculate the distribution by using five moments. This number
already provides almost as much information as the full frequency
distribution of the sample (see \sect~\ref{sec:marginalization}).
Figure~\ref{fig:plots_5mom} shows the resulting densities (that is,
distribution${}\times N$) for four example values of full-network sizes:
$N=n$, $N=1\,000$, $N=5\,000$, $N=10\,000$. The case $N=n$ corresponds to
applying the maximum-entropy method at the sample level; it can be observed
that with six moments it reproduces almost exactly the measured frequency
distribution. We discuss the case of unknown $N$ in \sect~\ref{sec:N}.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper_stensola_5mom_distributionsN_priorc.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper_riehle_5mom_distributionsN_priorc.pdf}%
\caption{***}
\label{fig:plots_5mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

The figure shows that the distribution for the full-network is less flat
than the measured frequency distribution for the sample; their difference
increases with $N$. Most remarkably, for the first data set the
distribution has two distinct low-activity modes. For the second data
the distribution presents a small shoulder on the right of the
mode.% , suggestive of two activity regimes. We frame no hypotheses about the
% biological cause of these two modes (they could stem from the presence of
% different kinds of cells or modules)
These features are clearly not present in the sample or in the
maximum-entropy distribution at the sample level. The application of the
probability calculus thus reveals interesting possible features of the full
network. \textcolor{white}{If you find this you can claim a postcard from
  us.}

\subsection*{Use with moment constraints}
\label{sec:use_moments}

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_twomom_distributions10000_priorc.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the full-network
distributions, for $N=10\,000$, obtained constraining the first two moments
(equivalent to constraining means and correlations) and the first four
moments. The frequency distribution of the sample is also shown for
comparison. In both data sets the two-moment constraint leads to a
distribution quite different from the four-moment constraint. For the first
data set in particular, one distribution is bimodal, the other unimodal.
This visual difference shows directly -- without the need to calculate
relative entropies, for example -- that the first two moments alone are
\emph{not} informationally sufficient for this specific data set.

It is instructive to contrast the conclusions about correlation sufficiency
reached through this full-network analysis, with those we would have
reached by applying the maximum-entropy method \emph{at the sample
  level}, as traditionally done. Such a comparison involves an important
methodological caveat, so we discuss it in full in the next section.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-21T16:46:09+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

This sampling relation is even more straightforward if instead of power
moments we use \emph{normalized factorial moments} \citep{potts1953}. The
$m$th normalized factorial moment of a distribution $\pxx(\ya)$ is
\begin{equation}
  \label{eq:def_factorial_moment_a}
  \E\Biggl[\binom{\ya}{m} \Biggr]\bigg/\binom{n}{m}
  \defd
  \sum_{\ya=0}^{n} %m!\,
  \displaystyle\binom{\ya}{m} \bigg/ \binom{n}{m}\; \pxx(\ya),
  \qquad 1\le m \le n,
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
that is, the expected number of distinct $m$-tuples of simultaneously
spiking neurons (within a time-bin's width), normalized by the maximum
possible number of distinct $m$-tuples. % For
% example, with $m=2$, if $a=4$ neurons spike in a network of $n=5$, we
% have $\tbinom{4}{2}=6$ distinct pairs of simultaneously spiking neurons,
% and the total number of distinct pairs is $\tbinom{5}{2}=10$. The
% normalized number of spiking pairs is therefore $6/10$.
Note that the first $m$ factorial moments together provide the same
information as the first $m$ power moments together, and vice versa: they
are linearly related because $\tbinom{\ya}{m}$ is a polynomial in $\ya$ of
degree $m$. % Specifying one set is therefore
% equivalent to specifying the other set.
But the normalized factorial moments have a convenient property for our
analysis: \emph{the first $n$ normalized factorial moments for the sample
  and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \begin{split}
  \E\Biggl[\binom{\ya}{m} \Biggr]\bigg/\binom{n}{m}
  &=
  \E\Biggl[\binom{\yA}{m} \Biggr]\bigg/\binom{N}{m}
  \qquad\text{or}\\
    \sum_{\ya=0}^{n} %m!\,
  \binom{\ya}{m}\bigg/  \binom{n}{m}\; \pxx(\ya)
&=    \sum_{\yA=0}^{N} %m!\,
  \binom{\yA}{m}\bigg/  \binom{N}{m}\; \pxxx(\yA),\qquad
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  1\le m \le n.
\end{split}
\end{equation}
% where $\pxxx(\yA)$ is our distribution of belief about the full-network
% activity, and $\pxx$

@@

Before applying the formula above to two concrete data sets we want to add
two remarks about the maximum-entropy method that are seldom made in the
neuroscientific literature. They are important for the interpretation of
the results and are further discussed in \sect~\ref{sec:assumptions}.
First, a maximum-entropy distribution like $\px(\yA \| M, N)$ is the
zeroth-order approximation (in the sense of Laplace's
approximation)\citep[\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000}
of four different distributions for the full network, which differ
numerically from one another in higher-order approximations:
\begin{enumerate}[label=(\roman*)]
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
The validity of the maximum-entropy approximation decreases as the ratio
$N/T$ increases. Second, the maximum-entropy
method % (based on the \emph{Shannon} relative
% entropy, as opposed to other entropy functions such as Burg's
% \citeyear{burg1975})
rests on some implicit assumptions about the probabilities for the long-run
frequency distribution of activities, besides the assumptions mentioned at
the end of the previous paragraph. So the often heard statement that it
gives \enquote{the maximally unbiased (or non-committal) distribution} must
be taken with a grain of salt. \citep[\mynote{move to
  sect}]{jaynes1986d_r1996,portamana2009,portamana2017}.

@@

In the next section we apply the method just described to the data sets
from two actual recordings and discuss the properties of the resulting
distributions.

@@

We apply the method just described to two data sets publicly available in
the literature:
\begin{itemize}[wide]
\item The \textbf{first data set}, from \textcite[rat
  14147]{stensolaetal2012}, consists of $n=65$ neurons (27 of which
  classified as grid cells) from rat Medial Entorhinal Cortex, recorded for
  about 20 minutes. Their spikes are binned into $T=417\,641$ bins of
  $3\;\textrm{ms}$ width.
\item The \textbf{second data set}, from
  \textcite{rostamietal2016_r2017}, %\texttt{data\_RS.npy}
  consists of $n=159$ neurons from macaque Motor Cortex, recorded for about
  15 minutes. Their spikes are binned into $T=300\,394$ bins of
  $3\;\textrm{ms}$ width.
\end{itemize}

@@

We first calculate the distribution by using five moments. This number
already provides almost as much information as the full frequency
distribution of the sample (see next section). Figure~\ref{fig:plots_5mom}
shows the resulting densities (distribution${}\times N$) for three example
values of full-network sizes: $N=1\,000$, $N=5\,000$, $N=10\,000$. The
frequency distribution of the sample activity is also shown, and in the
plot it is indistinguishable from the maximum-entropy distribution for
$N=n$, that is, applied at the sample level. We discuss the case of unknown
$N$ in \sect~\ref{sec:N}.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper_stensola_5mom_distributionsN_priorc.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper_riehle_5mom_distributionsN_priorc.pdf}%
\caption{***}
\label{fig:plots_5mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb

The figure shows that the distribution for the full-network is less flat
than the measured frequency distribution for the sample; their difference
increases with $N$. Most remarkably, for the first data set the
distribution has two distinct low-activity modes. For the second data set
the distribution presents a small shoulder on the right of the
mode.% , suggestive of two activity regimes. We frame no hypotheses about the
% biological cause of these two modes (they could stem from the presence of
% different kinds of cells or modules)
These features are clearly not present in the sample frequencies or in the
maximum-entropy distribution at the sample level. The application of the
probability calculus thus reveals interesting possible features of the full
network. \textcolor{white}{If you find this you can claim a postcard from
  us.}

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_twomom_distributions10000_priorc.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
To illustrate how our approach can be applied to studies of
\enquote{cooperativity} or \enquote{interaction},
\fig~\ref{fig:plots_2_4mom} shows the full-network distributions, for
$N=10\,000$, obtained constraining the first two moments (in
\textcolor{red}{red}, equivalent to constraining means and correlations)
and the first four moments (in \textcolor{blue}{blue}). The frequency
distribution for the sample is also shown for comparison. In both data sets
the two-moment constraint leads to a distribution quite different from the
that of the four-moment constraint. For the first data set in particular,
one distribution is bimodal, the other unimodal. Interpreted in terms of
This visual difference shows directly that the two sets of moments are not
even approximately equivalent.

It is instructive to contrast this conclusion with that we would have
reached by applying the maximum-entropy method \emph{at the sample level},
as traditionally done. Such a comparison involves an important
methodological caveat, so we discuss it in full in the next section.
